---
title: "End-to-end DADA2 pipeline with SpeedDada"
author: "Alexandre Jousset, Genedance GmbH"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{End-to-end DADA2 pipeline with SpeedDada}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "figures/"
)
```

# Introduction

`SpeedDada` provides a Rust-backed drop-in replacement for the
[dada2](https://github.com/benjjneb/dada2) amplicon sequence variant (ASV)
pipeline (Callahan et al. 2016, doi:10.1038/nmeth.3869). The seven exported
functions mirror dada2's R API and return objects of the same shapes, so
existing scripts run unchanged.

This vignette walks through the full paired-end pipeline using a tiny
FASTQ fixture shipped inside the package, so the example runs in seconds
on any platform.

# Setup

```{r load}
library(SpeedDada)

fwd <- system.file("extdata", "sample_R1.fastq", package = "SpeedDada")
rev <- system.file("extdata", "sample_R2.fastq", package = "SpeedDada")
fwd
```

# 1. Quality filtering

```{r filter}
fwd_filt <- tempfile(fileext = ".fastq")
rev_filt <- tempfile(fileext = ".fastq")
stats <- filterAndTrim(
  fwd, fwd_filt, rev, rev_filt,
  truncLen = c(30, 30),
  maxEE    = c(5, 5),
  minLen   = 10
)
stats
```

# 2. Learn the error model

```{r learn-errors}
err <- learnErrors(c(fwd_filt, rev_filt), nbases = 1e4)
```

# 3. Dereplicate and denoise

```{r dada}
dF <- derepFastq(fwd_filt)
dR <- derepFastq(rev_filt)

aF <- dada(dF, err, omega_a = 1e-5)
aR <- dada(dR, err, omega_a = 1e-5)
head(aF$denoised)
```

# 4. Merge pairs and build the sequence table

```{r merge}
merged <- mergePairs(aF, dF, aR, dR, minOverlap = 5)
head(merged)

seqtab <- makeSequenceTable(list(s1 = merged))
dim(seqtab)
```

# 5. Remove chimeras

```{r chimera}
seqtab_clean <- removeBimeraDenovo(seqtab)
dim(seqtab_clean)
```

# Cross-sample pooling

`SpeedDada` supports both `pool = TRUE` (full cross-sample pooling) and
`pool = "pseudo"` (Callahan's two-pass pseudo-pooling). Provide a *named
list* of `derep` objects as the first argument to `dada`:

```{r pool, eval = FALSE}
multi <- list(s1 = dF, s2 = dR)
res_pseudo <- dada(multi, err, pool = "pseudo")
```

# Session info

```{r sessioninfo}
sessionInfo()
```
